Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-24 submission/manuscript.Rmd
fix formatting discussion

- change first sentence: Other chromatin regulators involved in germline gene repression act in ESCs and EpiLCs. We observed this is true for KDM5C
We then globally characterized KDM5C binding at germline-enriched gene promoters through analysis of KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs). While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress germline transcripts[@scandagliaLossKdm5cCauses2017]. Although KDM5C is enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of EpiLC germline-enriched DEGs. One notable DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells that promotes meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Retinoic acid can only activate _Stra8_ expression when DAZL is present, which is a direct target of KDM5C and is aberrantly transcribed and translated in _Kdm5c_-KO EpiLCs. This indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells indirectly of KDM5C regulation through activation by other ectopic germline programs. Consistent with this idea, many _Kdm5c_-KO EpiLC germline DEGs are important for early germ cell development and meiosis while those expressed in the mature _Kdm5c_-KO brain are involved in late sperm development. Altogether, this indicates ectopic germline programs are, to some extent, progressing through germ cell developmental stages over the course of _Kdm5c_-KO development.

@@

It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, although primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, _Dazl_ is also expressed in na&#239;ve ESCs _in vitro_[@wellingDAZLRegulatesTet12015], the inner cell mass _in vivo_[@wellingDAZLRegulatesTet12015], and at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], but is silenced when cells transition from na&#239ve to primed pluripotency during ESC to EpiLC differentiation[@mochizukiRepressionGermlineGenes2021]. Very recently, two screens of _Dazl_-repressors in ESCs identified KDM5C as a direct regulator of _Dazl_ expression[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. Interestingly, one screen found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency in ESCs[@guptaGenomewideScreenReveals2023]. We found that while KDM5C also repressess _Dazl_ expression in EpiLCs, _Kdm5c_-KO EpiLCs do not express 2-cell specific genes like _Dux_ and _Zscan4c_. Out of the four 2-cell regulators characterized in ESCs, KDM5C was the only factor whose repression of _Dazl_ was independent of _Dux_ expression[@guptaGenomewideScreenReveals2023]. Together, shis suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression, including germline genes that are also involved in pluripotency and self-renewal.

@@

- We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020].

>>>>

- change first sentence: Other chromatin regulators involved in germline gene repression act in ESCs and EpiLCs. We observed this is true for KDM5C

We then globally characterized KDM5C binding at germline-enriched gene promoters through analysis of KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs). While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress germline transcripts[@scandagliaLossKdm5cCauses2017]. Although KDM5C is enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of EpiLC germline-enriched DEGs. One notable DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells that promotes meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Retinoic acid can only activate _Stra8_ expression when DAZL is present, which is a direct target of KDM5C and is aberrantly transcribed and translated in _Kdm5c_-KO EpiLCs. This indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells indirectly of KDM5C regulation through activation by other ectopic germline programs. Consistent with this idea, many _Kdm5c_-KO EpiLC germline DEGs are important for early germ cell development and meiosis while those expressed in the mature _Kdm5c_-KO brain are involved in late sperm development. Altogether, this indicates ectopic germline programs are, to some extent, progressing through germ cell developmental stages over the course of _Kdm5c_-KO development.

@@

It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, although primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, _Dazl_ is also expressed in na&#239;ve ESCs _in vitro_[@wellingDAZLRegulatesTet12015], the inner cell mass _in vivo_[@wellingDAZLRegulatesTet12015], and at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], but is silenced when cells transition from na&#239;ve to primed pluripotency during ESC to EpiLC differentiation[@mochizukiRepressionGermlineGenes2021]. Very recently, two screens of _Dazl_-repressors in ESCs identified KDM5C as a direct regulator of _Dazl_ expression[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. Interestingly, one screen found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency in ESCs[@guptaGenomewideScreenReveals2023]. We found that while KDM5C also repressess _Dazl_ expression in EpiLCs, _Kdm5c_-KO EpiLCs do not express 2-cell specific genes like _Dux_ and _Zscan4c_. Out of the four 2-cell regulators characterized in ESCs, KDM5C was the only factor whose repression of _Dazl_ was independent of _Dux_ expression[@guptaGenomewideScreenReveals2023]. Together, shis suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression, including germline genes that are also involved in pluripotency and self-renewal.

@@

- We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020]. **note: maybe just put in results**. Could move around depending upon if I get pheno working.
